Developmental trajectories of word recognition: a priming study in bilinguals

### 2023-11-03 manuscript/manuscript.qmd
Add results and exploratory analysis

targets::tar_load(
	c(participants,
	  stimuli,
	  vocabulary,
	  attrition_trials,
	  attrition_participants,
	  bvq_data,
	  gaze,
	  data_time_bcn
	))

@@

Building a mental lexicon is a major achievement in the development of an infant. By storing representations of how familiar words sound and what they mean, an infant is able to make sense of their linguistic input. Infants start forming their first lexical representations before their first year of life [@jusczyk1999; @halle1994; @vihman2004; @parise2012electrophysiological; @bergelson20126; @bergelson2015early]. This initial lexicon consists of only a few items; mainly words for people, interjections, body parts, and food [@tincoff2012six; @tardif2008baby], but it undergoes rapid growth during their second year of life [@ganger2004reexamining; @goldfield1990early; @bloom2002children; @mcmurray2007defusing]. According to parental reports the average 15-month-old infant already understands more than 100 words, and by two years, they understand more than 400 words [@frank2021variability]. This accelerated lexical developmental is reflected in infants' trajectories of word recognition: infants recognise familiar words faster and more efficiently as they approach their second birthday [@fernald1998rapid; @fernald2001half; @hurtado2007spoken].

Despite being exposed to a more complex linguistic input, bilinguals show equivalent trajectories of word acquisition and word recognition to their monolingual peers' [@vihman2007onset; @vihman2013; @legacy2018vocabulary; @hoff2012; @de2014bilingual; @pearson1994patterns; @byers-heinlein2023sometimes]. This is a remarkable deed for two reasons. First, bilingual infants receive a relative impoverished linguistic input in each of their languages compared to monolinguals [@costa2014does]. Second, they face a more complex referential context, since they often learn two labels for each referent, one in each language. The mechanisms that allow bilingual infants to keep up with monolinguals in their lexical development trajectories are still unclear.

@@

More recently, @chow2017spoken adapted the Visual World Paradigm from @huettig2007tug to explore 24- to 30-month-old toddlers' visual fixation patterns during a word recognition when presented with phonological and semantic distractors. In each trial, the authors presented participants with four semantically and phonologically pictures. Four seconds after pictures onset, a word-form was auditorily presented. The word did not refer to any of the pictures displayed on the screen, but was phonologically related to one of them (both labels shared phonological onset), and semantically related to another one of the pictures (both referents belonged to the same taxonomic category. For instance, participants might be presented with the pictures of a sandwich, a bus, a cat, and a dress. Then they would hear the carrier phrase "Look at the **bee**!". The authors registered participants fixations to the phonological and semantic distractors, and found evidence of a preference for the phonological distractor at earlier stages of the post-naming phase, and a preference for the semantic distractor at later stages of the trial. These results support a cascaded activation account of lexical access during the first stages of lexical development, paralleling previous findings in adults [@allopenna1998tracking].

Bilinguals provide one of the most interesting pieces of evidence in support of a cascaded account of lexical processing. A critical property the bilingual lexicon is that its representations are accessed in a language-non selective way. Even during monolingual situations, bilinguals access their lexical representations in both of their languages [e.g.; @thierry2007brain; @dunabeitia2009masked]. This is reflected in bilinguals' performance in word recognition and production tasks. For instance, @marian1999activation presented Russian-English bilinguals with a series of trials that started with an auditory instruction like "Poloji marku nije krestika" ["Put the stamp below the cross"]. The instruction would be present in Russian (dominant language for most participants). The target object (*marku* [*stamp*], in this case) would be present in the screen, together with an object whose English label shared phonological onset with the target label (e.g., *mark*), and two unrelated distractors whose labels in Russian and English did not shared phonological relationship with the target label. Converging with @allopenna1998tracking's results in monolingual adults, after hearing the target label bilinguals fixated the cross-language phonological distractor significantly more than the unrelated distractors. These findings suggest that participants activated phonologically related word-forms in both Russian *and* English, which affected their overt visual exploration patterns during the task.

@@

Phonological priming paradigms have also revealed language non-selectivity in the initial bilingual lexicon. For instance, @von2012language found evidence of cross-language phonological priming in 21- to 42-months-old children learning German and English. At the beginning of each trial, the authors auditorily presented an English prime word embedded in a carrier phrase. Then, the target label was auditorily presented in German. Finally the target and distractor pictures were presented side-by-side. The authors manipulated the phonological overlap between the prime and target labels. The novelty in this study lied in the orthogonal manipulation of the phonological overlap between both labels in English and German. In some trials, the auditorily presented prime and target labels were phonologically related (e.g., *slide*-*Klide* [*dress*]). In some other trials, both labels were phonologically related through translation: the auditorily presented prime label did *not* overlap with the target label (e.g., *leg*-*Stein* [*stone*]), but its translation in German did [*Bein*]. In the rest of the trials prime and target labels were phonologically unrelated in both languages. The authors found a facilitation effect of cross-language priming when both prime and target auditory labels overlapped phonologically, as revealed by a stronger target picture looking preference, as compared to that of unrelated trials. Interestingly, participants showed a weaker target preference in priming through translation trials. Since participants' target recognition was interfered by the non-presented prime translation, such an inhibitory effect must stem at the lexical level. The fact that phonological priming facilitated the recognition of the target words (both labels belong to different languages), and that phonological priming through translation interfered target recognition (translation and target belong to the same language) points to cross-language connections playing an excitatory role, and to within-language connections playing an inhibitory role.

@floccia2020translation tested 27-month-old simultaneous bilinguals in a cross-language priming paradigm. The two languages participants were learning were English, and an Additional Language (either Cantonese, Dutch, French, German, Greek, Italian, Mandarin, Polish, Portuguese, or Spanish). A prime word was embedded at the end of a carrier sentence that participants listened to. Then participants were presented with the auditory label of the target word, and two pictures were shown side-by-side, the target picture, and a distractor picture. In Study 1, the target word was the translation equivalent of the prime word (*cheese*-*fromage* for an English-French bilingual). In other trials, prime and target words were unrelated (e.g., *sock*-*fromage*). There was no phonological overlap at onset between prime and target or distractor labels in either of the two conditions. Participants were presented with two blocks of related and unrelated trials. In one of the blocks, participants were presented with English prime words and Home Language target words (*sock*-*fromage*), and in the other block they were presented with Additional Language prime words and English target words (*fromage*-*sock*). The authors found a cross-linguistic priming effect, in which participants showed a stronger preference for the target picture in the post-naming phase in related trials, as compared to unrelated trials. This effect was found regardless of the language of the prime (English or Home Language), or the dominance of the language of the prime (i.e., whether the prime word belonged to the dominant language). Target preference exceeded chance level in both conditions. In Study 2, the authors tested a similar group of bilinguals in a cross-linguistic semantic priming task, in which participants were primed with words in English, and tested with semantically related words in the Additional Language (or vice versa). This time, prime and target words were not translation equivalents, but rather words whose referents shared semantic features. Results from Study  2 were parallel to those from Study 1. The authors found a strong semantic priming effect across languages, regardless of the language of the prime or the dominance of the language of the prime. Overall these results suggest that bilinguals recognised target words in both conditions, and benefited from a cross-linguistic facilitation effect from translation equivalents. The symmetry of such effect across both languages suggests that participants lexical access occurred in parallel for the two members of translation equivalents. As @floccia2020translation note, the stimuli set of Study 2 contains multiple cognates, especially in languages with a larger degree of phonological overlap across translation pairs, like English and Dutch. The authors run *post hoc* analyses exploring the effect of cognateness on the cross-language priming effects found, which showed no evidence of a cognateness effect on target preference. However, the cross-language effects reported in Study 2 might still reflect the confounding influence of cognateness, as the resulting effect might have been generated by an interaction by the phonological and semantic overlap between cognate primes and targets.

The aforementioned studies use priming-though-translation paradigms. In these paradigms, cross-linguistic priming results from activation spreading from the lexical representation of the the (auditorily presented) prime word to the lexical representation of its translation through semantic associations (e.g., translation equivalence). When prime words are auditorily presented, the task design constraints the naming context to one of the languages. The consequence is a potential asymmetry between the activation dynamics of the lexical representation of both labels: while the participant has heard one label, the activation of its translation must spread across semantic links. This might lead to a lower strength, and higher latency of the activation of the translation.

Priming through translation paradigms prove the existence of cross-language phonological associations at the lexical level, as suggested by the fact that non-presented prime translations can interfere with subsequent target recognition. However, since participants are auditorily presented with a prime label, the contribution of sub-lexical facilitation through acoustic similarity cannot be ruled out. For instance, it is possible that the inhibitory priming through translation found by @von2012language is the result

THERE IS NO EVIDENCE OF IMPLICIT NAMING IN BILINGUALS

@@

The present study is aimed at investigating the developmental trajectories of word recognition in bilingualism, and how (if at all) language non-selectivity shapes them. Using an adaptation of [@mani2010infants], we presented 20- to 32-months-old toddlers with a cross-linguistic phonological priming paradigm. We manipulated both the phonological overlap between the prime word and target word, and the cognate status of the prime word, to test the hypothesis that bilingual toddlers implicitly name familiar pictures in both languages. If so, phonological priming effect should be larger when the prime label in  both languages shares phonological overlap with the target words (cognate condition), compared to when only the label of the prime in one language overlaps phonologically with the target (non-cognate condition). We also also included a control *Unrelated* condition in which regardless of the cognate status of the prime word, neither of the prime labels in any language overlapped phonologically with the target word.

@@

We collected data from `r n_participants_total$n[1]` children living in the Metropolitan Area of Barcelona (Spain), tested at the Laboratori de Recerca en Infància at the Universitat Pompeu Fabra. Families were recruited from maternity rooms in private hospitals and social media, and contacted via phone when the child's age spanned between 20 and 32 months. From the `r n_participants_total$n` children that participated, `r n_participants_sessions$session_1` participated once, `r n_participants_sessions$session_2` participated twice, and `r n_participants_sessions$session_3` participated three times. Recurrent participants were tested with at least XXX months of difference. We gathered a total of `r n_sessions_total$n` testing sessions. Participants were divided into monolinguals and bilinguals based on their relative degree of exposure to Catalan and Spanish, estimated using the Language Exposure Questionnaire [LEQ, @bosch2001evidence]. We categorised participants as monolingual if exposed to more than 80% or more of the time to their dominant language, and as bilingual otherwise. `r Words::English(n_participants_lp$n[1])` of the participants were categorised as monolinguals (`r n_participants_lp_sex[1]` female, `r n_participants_lp_sex[2]` male) and `n_participants_lp$n[2]` as Catalan/Spanish bilinguals (`r n_participants_lp_sex[3]` female, `r n_participants_lp_sex[4]` male) (see @tbl-participants for a detailed summary of participants' age and language profile). Participants' vision was normal, none used glasses or any other type of vision corrector.

```{r tbl-participants}
#| label: tbl-participants
#| tbl-cap: "Demographic and linguistic profile of testing sessions."
participants_bcn |> 
	unnest_wider(doe) |> 
	select(-matches("others")) |> 
	arrange(lp, test_language) |> 
	summarise(n = n(),
			  n_participants = n_distinct(child_id),
			  across(age, lst(mean, sd, min, max)),
			  across(matches("doe_"), lst(mean, sd)),
			  age_vctr = list(age),
			  .by = c(lp, test_language)) |> 
	gt(rowname_col = "test_language",
	   groupname_col = "lp") |> 
	gt_plt_dist(age_vctr, type = "histogram") |>
	fmt_number(is.numeric, decimals = 1) |> 
	fmt_integer(c(n, n_participants)) |> 
	fmt_number(matches("doe"), decimals = 1, scale_by = 100) |> 
	cols_merge_range(age_min, age_max) |> 
	cols_merge_uncert(age_mean, age_sd) |> 
	cols_merge_uncert(doe_catalan_mean, doe_catalan_sd) |> 
	cols_merge_uncert(doe_english_mean, doe_english_sd) |> 
	cols_merge_uncert(doe_spanish_mean, doe_spanish_sd) |> 
	tab_spanner("Age (months)", matches("age")) |> 
	tab_spanner("Catalan", matches("catalan")) |> 
	tab_spanner("English", matches("english")) |> 
	tab_spanner("Spanish", matches("spanish")) |> 
	tab_spanner("Degree of Exposure (%)", matches("doe")) |> 
	cols_label(n = "N sessions",
			   n_participants = "N participants",
			   age_mean = "M (SD)",
			   age_min = "Range",
			   age_vctr = "Distribution",
			   doe_catalan_mean = "M (SD)",
			   doe_english_mean = "M (SD)",
			   doe_spanish_mean = "M (SD)") |> 
	tab_style(cell_text(style = "italic"),
			  cells_column_labels()) |> 
	tab_style(cell_text(weight = "bold"),
			  cells_column_spanners())
```

@@

```{r vocab-values}
vocabulary_bcn <- inner_join(vocabulary, 
							 select(participants_bcn, session_id),
							 by = join_by(session_id))

@@

We collected vocabulary data using parental responses to the Barcelona Vocabulary Inventory [BVQ, @garcia-castro2023bvq], an online vocabulary checklist inspired in several adaptations of the the Communicative Developmental Inventory [CDI, @fenson1994variability] developed to assess the vocabulary size of Catalan-Spanish bilingual toddlers. Families received a link to the BVQ immediately after each experimental session, and were given two weeks to fill it. We calculated two measures of receptive vocabulary size for each participant: *L1 vocabulary size* (proportion of words  reported as acquired in the checklist of the dominant language), and *total vocabulary size* (proportion of the words in both checklists reported as acquired) (see @fig-vocabulary).

`r n_imputed` (`r prop_imputed`) Families failed to provide a complete response to the BVQ within the two-week time limit, or did not provide a successful response to the questionnaire. For missing questionnaire responses, we imputed the vocabulary size of the participant using single imputation, using the vocabulary size scores of a pool of `r n_pool` additional participants for which a successful response for the questionnaire had been gathered. We used participants age in months and their language profile (monolingual/bilingual) as predictors. We used the `mice` R package [@van2011mice] to perform imputation using the Bayesian linear regression method.

```{r fig-vocabulary}
#| label: fig-vocabulary
#| fig-cap: "Participant vocabulary sizes across ages and language profiles. Vocabulary size scores are presented using two measures: L1 vocabulary size (proportion of words marked as *Understands* in the vocabulary checklist of the dominant language), and total vocabulary size (proportion of words marked as *Understands* in vocabulary checklists of both dominant and non-dominant languages). Vocabulary size scores from the recurrent participants are shown linked."
#| eval: false
#| fig-width: 8
#| fig-height: 4.5
vocabulary_bcn |> 
	inner_join(select(participants, child_id, session_id, age, lp),
			   by = join_by(child_id, session_id)) |> 
	pivot_longer(c(total_prop, l1_prop),
				 names_to = "measure",
				 values_to = "prop",
				 names_transform = \(x) gsub("_prop", "", x)) |> 
	mutate(measure = if_else(measure=="l1", 
							 "L1 vocabulary", 
							 "Total vocabulary"),
		   is_imputed = if_else(is_imputed, "Imputed", "Observed") |> 
		   	factor(levels = c("Observed", "Imputed"))) |> 
	ggplot(aes(age, prop, colour = is_imputed)) +
	facet_grid(measure ~ lp) +
	geom_line(aes(group = child_id),
			  alpha = 1/2,
			  colour = "grey") +
	geom_point() +
	labs(x = "Age (months)",
		 y = "Vocabulary size (%)") +
	theme(panel.grid.major = element_line(linetype = "dotted",
										  colour = "grey"),
		  legend.position = "top",
		  legend.title = element_blank())

@@

We examined the overall equivalence of the three trial types by comparing them across three variables relating to the target word: lexical frequency, word prevalence, animacy
@tbl-stimuli shows a detailed summary of the stimuli properties, broken down by trial type and testing language. Lexical frequencies were extracted from the Catalan and Spanish corpora of the CHILDES database [@reference; @reference] as counts per million words, and transformed into Zipf scores for easier cross-language comparison [@reference; @reference]. We defined word prevalence as the proportion of same-aged infants who were reported to understand the word in the BVQ database.

@@

For each word, we created a picture with a typical referent. To avoid competition between target and distractor pictures, semantically related target-distractor pairs were perceptually distinct [@floccia2020translation; @arias-trejo2010].

@@

### Data processing

@@

n_exc_prime <- sum(!attrition$is_valid_gaze_prime)
n_exc_test <- sum(!attrition$is_valid_gaze_test)
n_exc_test_any <- sum(!attrition$is_valid_gaze_test_any)
n_exc_participants <- sum(!attrition_participants$is_valid_participant)

@@

>>>>

targets::tar_load(
	c(participants,
	  stimuli,
	  vocabulary,
	  attrition_trials,
	  attrition_participants,
	  bvq_data,
	  gaze,
	  data_bcn
	))

@@

Building a mental lexicon is a major achievement in the development of an infant. By storing representations of how familiar words sound and what they mean, an infant is able to make sense of their linguistic input. Infants start forming their first lexical representations before their first year of life [@tincoff1999some; @halle1994emergence; @vihman2004cross; @parise2012electrophysiological; @bergelson20126; @bergelson2015early]. This initial lexicon consists of only a few items; mainly words for people, interjections, body parts, and food [@tincoff2012six; @tardif2008baby], but it undergoes rapid growth during their second year of life [@ganger2004reexamining; @goldfield1990early; @bloom2002children; @mcmurray2007defusing]. According to parental reports the average 15-month-old infant already understands more than 100 words, and by two years, they understand more than 400 words [@frank2021variability]. This accelerated lexical developmental is reflected in infants' trajectories of word recognition: infants recognise familiar words faster and more efficiently as they approach their second birthday [@fernald1998rapid; @fernald2001half; @hurtado2007spoken].

Despite being exposed to a more complex linguistic input, bilinguals show equivalent trajectories of word acquisition and word recognition to their monolingual peers' [@vihman2007onset; @legacy2018vocabulary; @hoff2012dual; @de2014bilingual; @pearson1994patterns; @byers-heinlein2023sometimes]. This is a remarkable deed for two reasons. First, bilingual infants receive a relative impoverished linguistic input in each of their languages compared to monolinguals [@costa2014does]. Second, they face a more complex referential context, since they often learn two labels for each referent, one in each language. The mechanisms that allow bilingual infants to keep up with monolinguals in their lexical development trajectories are still unclear.

@@

@chow2017spoken adapted the Visual World Paradigm from @huettig2007tug to explore 24- to 30-month-old toddlers' visual fixation patterns during a word recognition when presented with phonological and semantic distractors. In each trial, the authors presented participants with four semantically and phonologically pictures. Four seconds after pictures onset, a word-form was auditorily presented. The word did not refer to any of the pictures displayed on the screen, but was phonologically related to one of them (both labels shared phonological onset), and semantically related to another one of the pictures (both referents belonged to the same taxonomic category. For instance, participants might be presented with the pictures of a sandwich, a bus, a cat, and a dress. Then they would hear the carrier phrase "Look at the *bee*!". The authors registered participants fixations to the phonological and semantic distractors, and found evidence of a preference for the phonological distractor at earlier stages of the post-naming phase, and a preference for the semantic distractor at later stages of the trial. These results support a cascaded activation account of lexical access during the first stages of lexical development, paralleling previous findings in adults [@allopenna1998tracking].

One of the most interesting pieces of evidence in support of a cascaded account of lexical processing is provided by bilinguals. A critical property the bilingual lexicon is that its representations are accessed in a language-non selective way. Even during monolingual situations, bilinguals access their lexical representations in both of their languages [e.g., @thierry2007brain; @dunabeitia2009masked]. This is reflected in bilinguals' performance in word recognition and production tasks. For instance, @marian1999activation presented Russian-English bilinguals with a series of trials that started with an auditory instruction like "Poloji marku nije krestika" ["Put the stamp below the cross"]. The instruction would be present in Russian (dominant language for most participants). The target object (*marku* [*stamp*], in this case) would be present in the screen, together with an object whose English label shared phonological onset with the target label (e.g., *mark*), and two unrelated distractors whose labels in Russian and English did not shared phonological relationship with the target label. Converging with @allopenna1998tracking's results in monolingual adults, after hearing the target label bilinguals fixated the cross-language phonological distractor significantly more than the unrelated distractors. These findings suggest that participants activated phonologically related word-forms in both Russian *and* English, which affected their overt visual exploration patterns during the task.

@@

Phonological priming paradigms have also revealed language non-selectivity in the initial bilingual lexicon [@poarch2012cross; @bosma2020cognate; @singh2014one; @jardak2019labels; @von2019impact]. For instance, @von2012language found evidence of cross-language phonological priming in 21- to 42-months-old children learning German and English. At the beginning of each trial, the authors auditorily presented an English prime word embedded in a carrier phrase. Then, the target label was also auditorily presented in German. Finally the target and distractor pictures were presented side-by-side. The authors manipulated the phonological overlap between the prime and target labels. The novelty in this study lied in the orthogonal manipulation of the phonological overlap between both English and German labels. In some trials, the auditorily presented prime and target labels were phonologically related (e.g., *slide*-*Klide* [*dress*]). In some other trials, both labels were phonologically related through translation: the auditorily presented prime label did *not* overlap with the target label (e.g., *leg*-*Stein* [*stone*]), but its translation in German did [*Bein*]. In the rest of the trials prime and target labels were phonologically unrelated in both languages. The authors found a facilitation effect of cross-language priming when both prime and target auditory labels overlapped phonologically, as revealed by a stronger target picture looking preference, as compared to that of unrelated trials. Interestingly, participants showed a weaker target preference in priming through translation trials.

Using a similar task, @floccia2020translation provided conflicting results to those of @von2012language. The authors tested 27-month-old simultaneous bilinguals in a cross-language priming paradigm. The two languages participants were learning were English, and an Additional Language (either Cantonese, Dutch, French, German, Greek, Italian, Mandarin, Polish, Portuguese, or Spanish). A prime word was embedded at the end of a carrier sentence that participants listened to. Then, participants were presented with the auditory label of the target word, and two pictures were shown side-by-side, the target picture, and a distractor picture. In Study 1, the target word was the translation equivalent of the prime word (*cheese*-*fromage* for an English-French bilingual). In other trials, prime and target words were unrelated (e.g., *sock*-*fromage*). There was no phonological overlap at onset between prime and target or distractor labels in either of the two conditions. Participants were presented with two blocks of related and unrelated trials. In one of the blocks, participants were presented with English prime words and Home Language target words (*sock*-*fromage*), and in the other block they were presented with Additional Language prime words and English target words (*fromage*-*sock*). The authors found a cross-linguistic priming effect, in which participants showed a stronger preference for the target picture in the post-naming phase in related trials, as compared to unrelated trials. This effect was found regardless of the language of the prime (English or Home Language), or the dominance of the language of the prime (i.e., whether the prime word belonged to the dominant language). Target preference exceeded chance level in both conditions. In Study 2, the authors tested a similar group of bilinguals in a cross-linguistic semantic priming task, in which participants were primed with words in English, and tested with semantically related words in the Additional Language (or vice versa). This time, prime and target words were not translation equivalents, but rather words whose referents shared semantic features. Results from Study  2 were parallel to those from Study 1. The authors found a strong semantic priming effect across languages, regardless of the language of the prime or the dominance of the language of the prime. Overall these results suggest that bilinguals recognised target words in both conditions, and benefited from a cross-linguistic facilitation effect from translation equivalents. The symmetry of such effect across both languages suggests that participants lexical access occurred in parallel for the two members of translation equivalents. As @floccia2020translation note, the stimuli set of Study 2 contains multiple cognates, especially in languages with a larger degree of phonological overlap across translation pairs, like English and Dutch. The authors run *post hoc* analyses exploring the effect of cognateness on the cross-language priming effects found, which showed no evidence of a cognateness effect on target preference. However, the cross-language effects reported in Study 2 might still reflect the confounding influence of cognateness, as the resulting effect might have been generated by an interaction by the phonological and semantic overlap between cognate primes and targets.

Previous studies had reported similar cross-language priming effects, but as @von2012language note, the conclusion that such effects occurred at the lexical level, as opposed to being the result of mere acoustic similarity, was not warranted. In studies where the auditorily presented prime label was phonologically overlapped with the target label [@REFERENCE], priming effects might be explained by the activation of the phonological segments of the prime label, which facilitated the recognition of the target label, which shared such phonological segments. The priming-through-translation results by @von2012language and @floccia2020translation demonstrates that cross-language activation in the developing lexicon occurs at the lexical level. Since participants' target recognition was interfered by the non-presented prime translation, such an inhibitory effect must stem at the lexical level. The fact that phonological priming facilitated the recognition of the target words (both labels belong to different languages), and that phonological priming through translation interfered target recognition (translation and target belong to the same language) points to cross-language connections playing an excitatory role, and to within-language connections playing an inhibitory role.

Priming-though-translation paradigms, however, come with a cost. When prime words are auditorily presented, the task constrains the naming context to one of the languages. The consequence is a potential asymmetry between the activation dynamics of the lexical representation of both labels: while the participant has heard one label, the activation of its translation must spread across semantic links. This might lead to a lower strength, and higher latency of the activation of the translation.

@@

The present study is aimed at investigating the developmental trajectories of word recognition in bilingualism, and how (if at all) language non-selectivity shapes them. Using an adaptation of [@mani2010infant], we presented 20- to 32-months-old toddlers with a cross-linguistic phonological priming paradigm. We manipulated both the phonological overlap between the prime word and target word, and the cognate status of the prime word, to test the hypothesis that bilingual toddlers implicitly name familiar pictures in both languages. If so, phonological priming effect should be larger when the prime label in  both languages shares phonological overlap with the target words (cognate condition), compared to when only the label of the prime in one language overlaps phonologically with the target (non-cognate condition). We also also included a control *Unrelated* condition in which regardless of the cognate status of the prime word, neither of the prime labels in any language overlapped phonologically with the target word.

@@

We collected data from `r n_participants_total` children living in the Metropolitan Area of Barcelona (Spain), tested at the Laboratori de Recerca en Infància at the Universitat Pompeu Fabra. Families were recruited from maternity rooms in private hospitals and social media, and contacted via phone when the child's age spanned between 20 and 32 months. From the `r n_participants_total` children that participated, `r n_participants_sessions$session_1` participated once, `r n_participants_sessions$session_2` participated twice, and `r n_participants_sessions$session_3` participated three times. Recurrent participants were tested with at least XXX months of difference. We gathered a total of `r n_sessions_total$n` testing sessions. Participants were divided into monolinguals and bilinguals based on their relative degree of exposure to Catalan and Spanish, estimated using the Language Exposure Questionnaire [LEQ, @bosch2001evidence]. We categorised participants as monolingual if exposed to more than 80% or more of the time to their dominant language, and as bilingual otherwise. `r english::Words(n_participants_lp$n[1])` of the participants were categorised as monolinguals (`r n_participants_lp_sex$n[1]` female, `r n_participants_lp_sex$n[2]` male) and `r n_participants_lp$n[2]` as Catalan/Spanish bilinguals (`r n_participants_lp_sex$n[3]` female, `r n_participants_lp_sex$n[4]` male) (see @tbl-participants for a detailed summary of participants' age and language profile). Participants' vision was normal, none used glasses or any other type of vision corrector.

```{r tbl-participants}
#| label: tbl-participants
#| tbl-cap: "Demographic and linguistic profile of testing sessions."
participants_bcn |> 
	unnest_wider(doe) |> 
	select(-matches("others")) |> 
	arrange(lp, test_language) |> 
	summarise(n = n(),
			  n_participants = n_distinct(child_id),
			  across(age, lst(mean, sd, min, max)),
			  across(matches("doe_"), lst(mean, sd)),
			  # age_vctr = list(age),
			  .by = c(lp, test_language)) |> 
	gt(rowname_col = "test_language",
	   groupname_col = "lp") |> 
	# gt_plt_dist(age_vctr, type = "histogram") |>
	fmt_number(is.numeric, decimals = 1) |> 
	fmt_integer(c(n, n_participants)) |> 
	fmt_number(matches("doe"), decimals = 1, scale_by = 100) |> 
	cols_merge(c(age_min, age_max), 
			   pattern = "{1} ({2})") |> 
	cols_merge(c(age_mean, age_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(doe_catalan_mean, doe_catalan_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(doe_english_mean, doe_english_sd),
			   pattern = "{1} ({2})") |> 
	cols_merge(c(doe_spanish_mean, doe_spanish_sd), 
			   pattern = "{1} ({2})") |> 
	tab_spanner("Age (months)", matches("age")) |> 
	tab_spanner("Catalan", matches("catalan")) |> 
	tab_spanner("English", matches("english")) |> 
	tab_spanner("Spanish", matches("spanish")) |> 
	tab_spanner("Degree of Exposure (%)", matches("doe")) |> 
	cols_label(n = "N sessions",
			   n_participants = "N participants",
			   age_mean = "M (SD)",
			   age_min = "Range",
			   # age_vctr = "Distribution",
			   doe_catalan_mean = "M (SD)",
			   doe_english_mean = "M (SD)",
			   doe_spanish_mean = "M (SD)") |> 
	tab_style(cell_text(style = "italic"),
			  cells_column_labels()) |> 
	tab_style(cell_text(weight = "bold"),
			  cells_column_spanners())
```

@@

```{r vocab-values}
vocabulary_bcn <- vocabulary |> 
	inner_join(select(participants_bcn, session_id, lp, age_group),
			   by = join_by(session_id))

vocab_means <- vocabulary_bcn |> 
	summarise(across(c(total_prop, l1_prop), lst(mean, sd)),
			  n = n(),
			  .by = c(age_group, lp)) %>%
	mutate(across(matches("prop"), scales::percent)) %>%
	split(paste0("age_", gsub(" months", "", .$age_group))) |> 
	map(\(x) split(x, tolower(x$lp)))

@@

We collected vocabulary data using parental responses to the Barcelona Vocabulary Inventory [BVQ, @garcia-castro2023bvq], an online vocabulary checklist inspired in several adaptations of the the Communicative Developmental Inventory [CDI, @fenson1994variability] developed to assess the vocabulary size of Catalan-Spanish bilingual toddlers. Families received a link to the BVQ immediately after each experimental session, and were given two weeks to fill it. We calculated two measures of receptive vocabulary size for each participant: *L1 vocabulary size* (proportion of words  reported as acquired in the checklist of the dominant language), and *total vocabulary size* (proportion of the words in both checklists reported as acquired). `r n_imputed` (`r prop_imputed`) Families failed to provide a complete response to the BVQ within the two-week time limit, or did not provide a successful response to the questionnaire. For missing questionnaire responses, we imputed the vocabulary size of the participant using single imputation, using the vocabulary size scores of a pool of `r n_pool` additional participants for which a successful response for the questionnaire had been gathered. We used participants age in months and their language profile (monolingual/bilingual) as predictors. We used the `mice` R package [@van2011mice] to perform imputation using the Bayesian linear regression method.

Overall, there was a substantial increase in both total and L1 vocabulary sizes associated to age (see @fig-vocabulary). Monolinguals showed overall higher L1 vocabulary sizes than monolinguals: they were reported to understand an average of `r vocab_means$age_21$monolingual$l1_prop_mean` (*SD* = `r vocab_means$age_21$monolingual$l1_prop_sd`) of the words at around 21 months, `r vocab_means$age_25$monolingual$l1_prop_mean` (*SD* = `r vocab_means$age_25$monolingual$l1_prop_sd`) at 25 months, and `r vocab_means$age_30$monolingual$l1_prop_mean` (*SD* = `r vocab_means$age_30$monolingual$l1_prop_sd`) at 30 months, while bilinguals  were reported to understand an average of `r vocab_means$age_21$bilingual$l1_prop_mean` (*SD* = `r vocab_means$age_21$bilingual$l1_prop_sd`) of the words at around 21 months, `r vocab_means$age_25$bilingual$l1_prop_mean` (*SD* = `r vocab_means$age_25$bilingual$l1_prop_sd`) at 25 months, and `r vocab_means$age_30$bilingual$l1_prop_mean` (*SD* = `r vocab_means$age_30$bilingual$l1_prop_sd`) at 30 months. Total vocabulary sizes showed the opposite pattern, with monolinguals infants being reported to understand `r vocab_means$age_21$monolingual$total_prop_mean` (*SD* = `r vocab_means$age_21$monolingual$total_prop_sd`) of the words at around 21 months, `r vocab_means$age_25$monolingual$total_prop_mean` (*SD* = `r vocab_means$age_25$monolingual$total_prop_sd`) at 25 months, and `r vocab_means$age_30$monolingual$total_prop_mean` (*SD* = `r vocab_means$age_30$monolingual$total_prop_sd`) at 30 months, and bilinguals being reported to understand `r vocab_means$age_21$bilingual$total_prop_mean` (*SD* = `r vocab_means$age_21$bilingual$total_prop_sd`) of the words at around 21 months, `r vocab_means$age_25$bilingual$total_prop_mean` (*SD* = `r vocab_means$age_25$bilingual$total_prop_sd`) at 25 months, and `r vocab_means$age_30$bilingual$total_prop_mean` (*SD* = `r vocab_means$age_30$bilingual$total_prop_sd`) at 30 months.

```{r fig-vocabulary}
#| label: fig-vocabulary
#| fig-cap: "Participant vocabulary sizes across ages and language profiles. Vocabulary size scores are presented using two measures: L1 vocabulary size (proportion of words marked as *Understands* in the vocabulary checklist of the dominant language), and total vocabulary size (proportion of words marked as *Understands* in vocabulary checklists of both dominant and non-dominant languages). Grey lines connect vocabulary size scores from the recurrent participants. For visualisation purposes, participants were group into categorical age groups at 21, 25, 30 months, the ages around which most participants were tested."
#| fig-width: 8
#| fig-height: 4.5
vocabulary_bcn |> 
	pivot_longer(c(total_prop, l1_prop),
				 names_to = "measure",
				 values_to = "prop",
				 names_transform = \(x) gsub("_prop", "", x)) |> 
	mutate(measure = if_else(measure=="l1", 
							 "L1 vocabulary", 
							 "Total vocabulary"),
		   is_imputed = if_else(is_imputed, "Imputed", "Observed") |> 
		   	factor(levels = c("Observed", "Imputed"))) |> 
	ggplot(aes(age_group, prop,
			   colour = lp,
			   fill = lp)) +
	facet_wrap(~measure) +
	geom_line(aes(group = child_id),
			  colour = "grey",
			  alpha = 1/4, linewidth = 3/4) +
	geom_dots(aes(side = if_else(lp=="Monolingual", "left", "right")),
			  dotsize = 2,
			  scale = 0.5,
			  alpha = 3/4) +
	stat_summary(fun.data = mean_se,
				 geom = "errorbar",
				 linewidth = 3/4,
				 width = 0.15,
				 colour = "black",
				 show.legend = FALSE,
				 position = position_dodge(width = 1.15)) +
	stat_summary(fun = mean,
				 geom = "point",
				 size = 2,
				 colour = "black",
				 show.legend = FALSE,
				 position = position_dodge(width = 1.15)) +
	# geom_point(alpha = 1) +
	labs(x = "Age (months)",
		 y = "Vocabulary size",
		 colour = "Group",
		 fill = "Group") +
	scale_y_continuous(labels = scales::percent) +
	theme(panel.grid.major.y = element_line(linetype = "dotted",
											colour = "grey"),
		  legend.position = "top",
		  legend.title = element_blank())

@@

We examined the overall equivalence of the three trial types by comparing them across three variables relating to the target word: lexical frequency, word prevalence, animacy
@tbl-stimuli shows a detailed summary of the stimuli properties, broken down by trial type and testing language. Lexical frequencies were extracted from the Catalan and Spanish corpora of the CHILDES database [@macwhinney2000childes; @sanchez2019childes] as counts per million words, and transformed into Zipf scores for easier cross-language comparison [@zipf1945meaning; @van2014subtlex]. We defined word prevalence as the proportion of same-aged infants who were reported to understand the word in the BVQ database.

@@

For each word, we created a picture with a typical referent. To avoid competition between target and distractor pictures, semantically related target-distractor pairs were perceptually distinct [@floccia2020translation; @arias2009lexical].

@@

#### Data processing

@@

n_exc_prime <- sum(!attrition$is_valid_gaze_prime)
n_exc_test <- sum(!attrition$is_valid_gaze_test)
n_exc_test_each <- sum(!attrition$is_valid_gaze_test_each)
n_exc_participants <- sum(!attrition_participants$is_valid_participant)

@@

#### Modelling approach

```{r data-analysis}
